---
title: "R Intro (part 3)"
date: "`r Sys.Date()`"
output:
  html_document: 
    df_print: paged
    toc: yes
    toc_float: true
    toc_depth: 3
    code_folding: show
editor_options:
  chunk_output_type: inline
---


```{r "setup", echo=FALSE, message=FALSE, warning=FALSE}
# load all the libraries
library(tidyverse)
library(swirl)
library(testthat)
```

# 1. `Tidy data` chapter in R for Data Science book

[*R for Data Science*](http://r4ds.had.co.nz/) is an excellent book and resource for approaches to data science in R. I highly recommend bookmarking the link and using it as a reference whenever you want to learn more about the most up-to-date way of dealing with specific data manipulation and processing problems. For the `swirl()` lessons in this assignment, the chapter on **Tidy data** will provide some valuable background. Please read the following sections (no need to do the exercises, there will be similar questions in swirl) in the [**Tiday data** chapter](http://r4ds.had.co.nz/tidy-data.html#tidy-data-1):

 - 12.1 Introduction
 - 12.2 Tidy data
 - 12.3 Spreading and gathering
 - 12.4 Separating and uniting


# 2. Grouping and chaining in dplyr

In the `swirl()` course menu, select **Getting and Cleaning Data** and complete the second lesson **2: Grouping and Chaining with dplyr** and answer the following questions.

```{r, eval=FALSE}
swirl()
```

> A. Complete the following code chunk by adding another pipe command at the end to turn the dataset into a dplyr type data set (this is the same `cran` data used extensively in this lesson). 

Note that you will always have to escape swirl (`Esc`) to be able to run your own code.

```{r}
cran <- swirl:::swirl_courses_dir() %>% 
  file.path(
    "Getting_and_Cleaning_Data", 
    "Grouping_and_Chaining_with_dplyr", 
    "2014-07-08.csv"
  ) %>%
  read.csv(as.is = TRUE) # extend pipeline here


cran
```

> B. Reproduce the last pipeline of the lesson (Hint: to pass linting, be mindful
of the code style rule to put spaces around all so-called `infix` operators - i.e. 
` = `, ` + `, ` ^ `, ` / `, etc.).

```{r}
# write pipeline here
```

> C. Highlight the new functions you just learned about on your dplyr cheatsheet.

# 3. Tidying Data with tidyr

In the `swirl()` course menu, select **Getting and Cleaning Data** and complete the third lesson **3: Tidying Data with tidyr** and answer the following questions.

```{r, eval=FALSE}
swirl()
```

> A. The below pipeline reads the SAT dataset discussed at the end of this lesson. Try to extend the pipeline with the last set of commands from the lessen (including the `select`, `gather`, `separate`, `group_by` and `mutate` to calculate the `total` and `prop` columns). Exclude the `print` so the result is stored in the `sat` object rather than piped onward to `print`.

```{r}
sat <- swirl:::swirl_courses_dir() %>% 
  file.path(
    "Getting_and_Cleaning_Data", 
    "Tidying_Data_with_tidyr", 
    "sat13.csv"
  ) %>%
  read.csv(as.is = TRUE) # extend pipeline here
  

sat
```


> B. Highlight the new functions you just learned about on your tidyr cheatsheet.

# 4. Lint and knit/render

> Make sure your code passes linting (`Lint current file` command from the `Addins` menu - shortcut `Alt/Option`+`Shift`+`L`) and knits succesfull (`Knit to HTML` in the `Knit` menu - `Cmd`+`Shift`+`K`). 

# Automated checks (do not edit)

The following are automatic checks for Travis.

```{r "tests"}
expect_true(exists("cran"))
expect_is(cran, "tbl_df")
expect_true(exists("sat"))
expect_true(all(c("count", "total", "prop") %in% names(sat)))
```



