---
title: "R Intro (part 3)"
date: "`r Sys.Date()`"
output:
  html_document: 
    df_print: paged
    toc: yes
    toc_float: true
    toc_depth: 3
    code_folding: show
editor_options:
  chunk_output_type: inline
---


```{r "setup", echo=FALSE, message=FALSE, warning=FALSE}
# load all the libraries
library(tidyverse)
library(swirl)
library(testthat)
library(glue)
```

# 1. `Tidy data` chapter in R for Data Science book

[*R for Data Science*](http://r4ds.had.co.nz/) is an excellent book and resource for approaches to data science in R. I highly recommend bookmarking the link and using it as a reference whenever you want to learn more about the most up-to-date way of dealing with specific data manipulation and processing problems. For the `swirl()` lessons in this assignment, the chapter on **Tidy data** will provide some valuable background. Please read the following sections (no need to do the exercises, there will be similar questions in swirl) in the [**Tiday data** chapter](http://r4ds.had.co.nz/tidy-data.html#tidy-data-1):

 - 12.1 Introduction
 - 12.2 Tidy data
 - 12.3 Spreading and gathering
 - 12.4 Separating and uniting


# 2. Grouping and chaining in dplyr

In the `swirl()` course menu, select **Getting and Cleaning Data** and complete the second lesson **2: Grouping and Chaining with dplyr** and answer the following questions.

```{r, eval=FALSE}
swirl()
```

> A. Complete the following code chunk by adding another pipe command at the end to turn the dataset into a dplyr type data set (this is the same `cran` data used extensively in this lesson). 

Note that you will always have to escape swirl (`Esc`) to be able to run your own code.

```{r}
cran <- swirl:::swirl_courses_dir() %>% 
  file.path(
    "Getting_and_Cleaning_Data", 
    "Grouping_and_Chaining_with_dplyr", 
    "2014-07-08.csv"
  ) %>%
  read.csv(as.is = TRUE) # extend pipeline here


cran
```

> B. Reproduce the last pipeline of the lesson (Hint: to pass linting, be mindful
of the code style rule to put spaces around all so-called `infix` operators - i.e. 
` = `, ` + `, ` ^ `, ` / `, etc.).

```{r}
# write pipeline here
```

> C. Highlight the new functions you just learned about on your dplyr cheatsheet.

# 3. Tidying Data with tidyr

In the `swirl()` course menu, select **Getting and Cleaning Data** and complete the third lesson **3: Tidying Data with tidyr** and answer the following questions.

```{r, eval=FALSE}
swirl()
```

> A. The below pipeline reads the SAT dataset discussed at the end of this lesson. Try to extend the pipeline with the last set of commands from the lessen (including the `select`, `gather`, `separate`, `group_by` and `mutate` to calculate the `total` and `prop` columns). Exclude the `print` so the result is stored in the `sat` object rather than piped onward to `print`.

```{r}
sat <- swirl:::swirl_courses_dir() %>% 
  file.path(
    "Getting_and_Cleaning_Data", 
    "Tidying_Data_with_tidyr", 
    "sat13.csv"
  ) %>%
  read.csv(as.is = TRUE) # extend pipeline here
  

sat
```


> B. Highlight the new functions you just learned about on your tidyr cheatsheet.

# 4. Code checks

> The following are automated tests for Travis. Run the chunk to see if your code is complete and passes all the tests. The chunk will stop at the first test that does not pass. If you get an error, see if you can figure out what might be missing in your code (do not edit this chunk!). Once all tests run without errors, the file knits successfully (`Knit to HTML` in the `Knit` menu - `Cmd`+`Shift`+`K`), and your code passes linting (`Lint current file` command from the `Addins` menu - shortcut `Alt/Option`+`Shift`+`L`) without warnings, your assignment is ready for submission.

```{r "tests"}
msg <- function(q, ...) glue("Problem with question {q}: ", ..., ".")

expect_true(
  exists("cran"), 
  info = msg("2A", "'cran' object should exist but does not")
)

expect_is(
  cran, "tbl_df",
  info = msg("2A", "'cran' object is not a dplyr style data frame")
) 

expect_true(
  exists("sat"),
  info = msg("3A", "'sat' object should exist but does not")
)

expect_true(
  all(c("count", "total", "prop") %in% names(sat)),
  info = msg("3A", 
             "'sat' object does not have all the expected columns: ",
             "'count', 'total', 'prop'")
)
```