---
title: Before we start - R basics
author: adapted from Data Carpentry
output:
  html_document: 
    df_print: paged
    toc: yes
    toc_float: true
    toc_depth: 3
    code_folding: show
editor_options:
  chunk_output_type: inline
---

# RStudio

* Four segments: Console/Terminal/R Markdown/Markers (bottom left), Scripts (top
  left), Environment/History/Git (top right), Plots/Files/Help (bottom right)
* Code and workflow are more reproducible if we can document everything that we
  do.
* Our end goal is not just to "do stuff" but to do it in a way that anyone can
  easily and exactly replicate our workflow and results.
* See RStudio Cheatsheet for help and useful keyboard shortcuts

# Interacting with R

The basis of programming is that we write down instructions for the computer to
follow, and then we tell the computer to follow those instructions. We write, or
*code*, instructions in R because it is a common language that both the computer
and we can understand. We call the instructions *commands* and we tell the
computer to follow the instructions by *executing* (also called *running*) those
commands.

There are two main ways of interacting with R: using the console or by using
script files (plain text files that contain your code).

The console window (in RStudio, the bottom left panel) is the place where R is
waiting for you to tell it what to do, and where it will show the results of a
command that has been executed (exception: RMarkdown can show the results of 
entire code chunks inline in the document). You can type commands directly into 
the console and press `Enter` to execute those commands, but they will be 
forgotten when you close the session. It is better to enter the commands in the
chunk of an RMarkdown file, and save the file. This way, you have a complete 
record of what you did, you can easily show others how you did it and you can do
it again later on if needed. Rstudio allows you to execute commands directly
from the script editor by using the `Cmd-Enter` shortcut for a single 
command (you don't need to select it and it can stretch across multipled lines, 
just have the cursor anywhere in the command) or the `Cmd-Option-C` to execute
an entire code chunk. 

At some point in your analysis you may want to check the content of variable or
the structure of an object, without necessarily keep a record of it in your
script. You can type these commands and execute them directly in the console.
RStudio provides the `Ctrl-1` and `Ctrl-2` shortcuts allow you to jump between
the script and the console windows.

If R is ready to accept commands, the R console shows a `>` prompt. If it
receives a command (by typing, copy-pasting or sent from the script editor using
`Cmd-Enter` or `Cmd-Option/Alt-C`), R will try to execute it, and when ready, show 
the results and come back with a new `>`-prompt to wait for new commands.

If R is still waiting for you to enter more data because it isn't complete yet,
the console will show a `+` prompt. It means that you haven't finished entering
a complete command. This is because you have not 'closed' a parenthesis or
quotation, i.e. you don't have the same number of left-parentheses as
right-parentheses, or the same number of opening and closing quotation marks. If
you're in Rstudio and this happens, click inside the console window and press
`Esc`; this will cancel the incomplete command and return you to the `>` prompt.
Linting a document (`Addins`->`Lint current file` or `Cmd-Option/Alt-L` will also
highlight coding errors like this in your RMarkdown file).

# Commenting

Use `#` signs to comment. Comment liberally in your R scripts. Anything to the
right of a `#` is ignored by R, meaning it won't be executed.

# Assignment operator

`<-` is the assignment operator. It assigns values on the right to objects on
the left. So, after executing `x <- 3`, the value of `x` is `3`. The arrow can
be read as 3 **goes into** `x`.  For historical reasons, you can also use ` = ` 
for assignments, but not in every context. Because of this, it is good practice 
to use always `<-` for assignments, except when specifying the values of
arguments in functions, when only ` = ` should be used, see below.

Inside a code chunk, the keyboard shortcut `Option/Alt`+`-` will write ` <- ` 
in a single keystroke.

# Functions and their arguments

Functions are "canned scripts" that automate something complicated or convenient
or both.  Many functions are predefined, or can be made available by importing R
*packages* (more on that later). A function usually gets one or more inputs
called *arguments*. Functions often (but not always) return a *value*. A typical
example would be the function `sqrt()`. The input (the argument) must be a
number, and the return value (in fact, the output) is the square root of that
number. Executing a function ('running it') is called *calling* the function. An
example of a function call is:

`b <- sqrt(a)`

Here, the value of `a` is given to the `sqrt()` function, the `sqrt()` function
calculates the square root, and returns the value which is then assigned to
variable `b`. This function is very simple, because it takes just one argument.

The return 'value' of a function need not be numerical (like that of `sqrt()`),
and it also does not need to be a single item: it can be a set of things, or
even a data set. We'll see that when we read data files in to R.

Arguments can be anything, not only numbers or filenames, but also other
objects. Exactly what each argument means differs per function, and must be
looked up in the documentation (see below). Some functions take arguments which
may either be specified by the user, or if left out, take on a *default* value:
these are called *options*. Options are typically used to alter the way the
function operates, such as whether it ignores 'bad values', or what symbol to
use in a plot.  However, if you want something specific, you can specify a value
of your choice which will be used instead of the default.

Let's try a function that can take multiple arguments: `round`.

```{r}
round(3.14159)
```

Here, we've called `round` with just one argument, `3.14159`, and it has
returned the value `3`.  That's because the default is to round to the nearest
whole number. If we want more digits we can see how to do that by getting
information about the `round` function.  We can use `args(round)` or look at the
help for this function using `?round`.

```{r}
args(round)
```

```{r, eval = FALSE}
?round
```

We see that if we want a different number of digits, we can
type `digits = 2` or however many we want.

```{r}
round(3.14159, digits = 2)
```

If you provide the arguments in the exact same order as they are defined you
don't have to name them:

```{r}
round(3.14159, 2)
```

And if you do name the arguments, you can switch their order:

```{r}
round(digits = 2, x = 3.14159)
```

It's good practice to put the non-optional arguments (like the number you're
rounding) first in your function call, and to specify the names of all optional
arguments.  If you don't, someone reading your code might have to look up
definition of a function with unfamiliar arguments to understand what you're
doing.

# Packages/libraries/modules

Extensions of the basic R are called *packages* (sometimes modules or libraries) 
and loaded via the `library(package_name)` function. There are thousands of 
packages with specific functionality. Many packages are available from the 
central R package distribution center CRAN (the [*Comprehensive R Archive Network*](https://cran.r-project.org/)) and can be installed from there using 
the `install.packages(...)` function or the *Package Manger* on the lower right 
corner of RStudio (Next to the *Help* tab). The package manager makes it easy to 
see what all is installed and which packages are currently loaded. Feel free to 
install additional packages (they will be only in your own user account) if you 
want to test how they work. Additionally, many more packages are available 
directly on GitHub. These can be installed with the `devtools::install_github()`
command.

# The pipe operator

The following operator `%>%` (sometimes called the **pipe** operator) is a very 
useful alternative to passing arguments to functions. The way it works is that 
whatever is to the left of this operator is sent (also called "piped") to the 
function that comes after operator (in the direction the `>` sign within `%>%` 
points) as its first argument. All additional arguments are specified as usual. 
We will revisit the example of using the function `round` above to illustrate 
how `%>%` works.

```{r}
library(magrittr) # load the magrittr library to include the pipe operator

# the following is the classic way ("nested") where each parameter 
# is inside the function
round(3.14159, digits = 2)

# this is using the pipe to send what's left of it (3.14159) to what comes 
# after (round). note that the digits parameter is still defined as usual.
3.14159 %>% round(digits = 2)
```

This operator might seem unnecessarily complicated in a simple one step 
operation such as the one listed above (and one would rarely use it for a 
single function call) but it becomes immensly powerful for chaining together 
multiple commands and keeping the whole sequence of operations well-documented 
and easy to follow. It is good practice to use `%>%` for more complex 
operations so it is good to get comfortable with how it works. As a simple 
example for chaining together 3 functions, take the following (a bit contrived)
math example:

```{r}
# the following are two identical approachs for taking the log (base 5) of a 
# number, calculating its square root and then rounding the result

# nested commands
round(sqrt(log(42, base = 5)), digits = 3)

# pipe 
42 %>% log(base = 5) %>% sqrt() %>% round(digits = 3)

# or across multiple lines with detailed documentation
42 %>% 
  log(base = 5) %>% # compute the logarithm in base 5
  sqrt() %>% # calculate the square root of the result
  round(digits = 3) # round the result
```


# Creating objects

You can get output from R simply by typing in math in the console

```{r}
3 + 5
12 / 7
```

However, to do useful and interesting things, we need to assign _values_ to
_objects_. To create an object, we need to give it a name followed by the
assignment operator `<-`, and the value we want to give it:

```{r}
weight_kg <- 55
```

Objects can be given any name such as `x`, `current_temperature`, or
`subject_id`. You want your object names to be explicit and not too long. They
cannot start with a number (`2x` is not valid, but `x2` is). R is case sensitive
(e.g., `weight_kg` is different from `Weight_kg`). There are some names that
cannot be used because they are the names of fundamental functions in R (e.g.,
`if`, `else`, `for`, `c`). It's also best to avoid dots (`.`) within a variable 
name as in `my.dataset`. There are many functions in R with dots in their names 
for historical reasons, but because dots have a special meaning in R (for 
methods) and other programming languages, it's best to avoid them. 

It is also recommended to use nouns for variable names, and verbs for function 
names. It's important to be consistent in the styling of your code
(where you put spaces, how you name variable, etc.). In R, two popular style
guides are [Hadley Wickham's](http://adv-r.had.co.nz/Style.html) and
[Google's](https://google-styleguide.googlecode.com/svn/trunk/Rguide.xml) with
the the prior used as the default style for linting.

Note that traditional variable names can only contain letters, numbers, the dot 
(`.`) and the underscore (`_`) and may not start with a number. However, if you 
reallyneed or want to use a variable name that deviates from these rules you can 
do  so but always have to type them with tickmarks ``` around the variables. 
Examples:

```{r}
# these are all valid traditional names
weight <- 65
weight_kg <- 67
weight2 <- 68

# these do not conform to the standard rules and have to go into the tickmarks
# (not recommended to use but okay if really needed/imported from a spreadsheet)
`height [cm]` <- 165 # contains a space and the special characters [ and ]
`2nd_height` <- 170 # variable starts with a number
```

## Exercise 1

> Write an expression to calculate the ratio of *weight_kg* to *height [cm]* 
using the variables defined above.

```{r}

```

# Printing and overwriting objects

When assigning a value to an object, R does not print anything. You can force to
print the value by explicitly printing it or by typing the name:

```{r}
weight_kg <- 55     # doesn't print anything
(weight_kg <- 55)   # putting parenthesis around the call prints the value
weight_kg           # and so does typing the name of the object
print(weight_kg)    # or explicitly printing the variable content
```

Now that R has `weight_kg` in memory, we can do arithmetic with it. For
instance, we may want to convert this weight in pounds:

```{r}
2.2 * weight_kg
```

We can also change a variable's value by assigning it a new one:

```{r}
weight_kg <- 57.5
2.2 * weight_kg
```

This means that assigning a value to one variable does not change the values of
other variables.  For example, let's store the weight in pounds in a new
variable, `weight_lb`:

```{r}
weight_lb <- 2.2 * weight_kg
```

and then change `weight_kg` to 100.

```{r}
weight_kg <- 100
```

What do you think is the current content of the object `weight_lb`? 126.5 or 200?

## Exercise 2

> What are the values after each statement in the following?

```{r}
mass <- 47.5              # mass?
age  <- 122               # age?
mass <- mass * 2.0        # mass?
age  <- age - 20          # age?
mass_index <- mass / age  # mass_index?
```

# Vectors and data types

A vector is the most common and basic data structure in R, and is pretty much
the workhorse of R. It's a group of values, mainly either numbers or
characters. You can assign this list of values to a variable, just like you
would for one item. For example we can create a vector of weights:

```{r}
weights <- c(50, 60, 65, 82)
weights
```

A vector can also contain characters:

```{r}
animals <- c("mouse", "rat", "dog")
animals
```

There are many functions that allow you to inspect the content of a
vector. `length()` tells you how many elements are in a particular vector:

```{r}
length(weights)
length(animals)
```

An important feature of a vector, is that all of the elements are the same type of data.
The function `class()` indicates the class (the type of element) of an object:

```{r}
class(weights)
class(animals)
```

The function `str()` provides an overview of the object and the elements it
contains. It is a really useful function when working with large and complex
objects:

```{r}
str(weights)
str(animals)
```

You can add elements to your vector by using the `c()` function:

```{r}
weights <- c(weights, 90) # adding at the end of the vector
weights <- c(30, weights) # adding at the beginning of the vector
weights
```

What happens here is that we take the original vector `weights`, and we are
adding another item first to the end of the other ones, and then another item at
the beginning. We can do this over and over again to grow a vector, or assemble
a dataset. As we program, this may be useful to add results that we are
collecting or calculating.

We just saw 2 of the 6 **atomic vector** types that R uses: `"character"` and
`"numeric"`. These are the basic building blocks that all R objects are built
from. The other 4 are:

* `"logical"` for `TRUE` and `FALSE` (the boolean data type)
* `"integer"` for integer numbers (e.g., `2L`, the `L` indicates to R that it's an integer)
* `"complex"` to represent complex numbers with real and imaginary parts (e.g.,
  `1+4i`) and that's all we're going to say about them
* `"raw"` that we won't discuss further (it's for binary data)

Vectors are one of the many **data structures** that R uses. Other important
ones are lists (`list`), matrices (`matrix`), data frames (`data.frame`) and
factors (`factor`).

## Exercise 3

We’ve seen that atomic vectors can be of type character, numeric, integer,
and logical. But what happens if we try to mix these types in a single
vector?

> What will happen in each of the following examples? (hint: use `class()` to 
check the data type of your object)


```{r}
num_char <- c(1, 2, 3, "a")

num_logical <- c(1, 2, 3, TRUE)

char_logical <- c("a", "b", "c", TRUE)

tricky <- c(1, 2, 3, "4")

```

> Why do you think it happens?


> Can you draw a diagram that represents the hierarchy of the data types?



# Missing data

As R was designed to work with data, it includes the concept of missing data
(which is uncommon in other programming languages). Missing data are represented
as `NA`.

```{r}
planets <- c("Mercury", "Venus", "Earth", "Mars", "Jupiter", "Saturn", "Uranus",
             "Neptune", NA)
```

When doing operations on numbers, most functions will return `NA` if the data
you are working with include missing values. It is a safer behavior as otherwise
you may overlook that you are dealing with missing data. You can add the
argument `na.rm = TRUE` to calculate the result while ignoring the missing values.

```{r}
heights <- c(2, 4, 4, NA, 6)
mean(heights)
max(heights)
mean(heights, na.rm  =  TRUE)
max(heights, na.rm  =  TRUE)
```

If your data include missing values, you may want to become familiar with the
functions `is.na()` and `na.omit()`.

# Seeking help

## I know the name of the function I want to use, but I'm not sure how to use it

If you need help with a specific function, let's say `barplot()`, you can type:

```{r, eval = FALSE}
?barplot
```

## I want to use a function that does X, there must be a function for it but I don't know which one...

If you are looking for a function to do a particular task, you can use
`help.search()` function, which is called by the double question mark `??`.
However, this only looks through the installed packages for help pages with a
match to your search request

```{r, eval = FALSE}
??kruskal
```

If you can't find what you are looking for, you can use the
[rdocumention.org](http://www.rdocumentation.org) website that search through
the help files across all packages available.

## I am stuck... I get an error message that I don't understand

Start by googling the error message. However, this doesn't always work very well
because often, package developers rely on the error catching provided by R. You
end up with general error messages that might not be very helpful to diagnose a
problem (e.g. "subscript out of bounds"). If the message is very generic, you might
also include the name of the function or package you're using in your query.

However, you should check StackOverflow. Search using the `[r]` tag. Most
questions have already been answered, but the Exercise is to use the right
words in the search to find the answers:
[http://stackoverflow.com/questions/tagged/r](http://stackoverflow.com/questions/tagged/r)

The [Introduction to R](http://cran.r-project.org/doc/manuals/R-intro.pdf) can
also be dense for people with little programming experience but it is a good
place to understand the underpinnings of the R language.

The [R FAQ](http://cran.r-project.org/doc/FAQ/R-FAQ.html) is dense and technical
but it is full of useful information.

## More resources


* [different R cheatsheets](https://www.rstudio.com/resources/cheatsheets/)
* [R for Data Science online book](http://r4ds.had.co.nz/)
* [RMarkdown code chunk options](http://yihui.name/knitr/options/)
* The [Posting Guide](http://www.r-project.org/posting-guide.html) for the R
  mailing lists.
* [How to ask for R help](http://blog.revolutionanalytics.com/2014/01/how-to-ask-for-r-help.html)
  useful guidelines
* [This blog post by Jon Skeet](http://codeblog.jonskeet.uk/2010/08/29/writing-the-perfect-question/)
  has quite comprehensive advice on how to ask programming questions.
